# Tests of Mean Differences

The test of a mean difference is one of the principal tools for drawing inferences from experimental designs. Random assignment cancels out individual differences and background factors, allowing the *average causal effect* to be assessed. Generalising the result still calls on the machinery of inferential statistics, and considerations of sample size and Type I/II error remain.

## The one-sample test

We begin with the **one-sample test**. It applies when one wishes to judge whether a sample mean is statistically distinguishable from some specified value — a known population mean, or a value posited by theory. For example, having collected responses on a 7-point Likert scale, one may ask whether the mean of an item is meaningfully different from the scale midpoint (4). Suppose we have ten Likert responses; we represent them by drawing ten normal variates with mean 4 and SD 1. In actual research these values would be observed responses on the scale.

```{r,oneTsample}
pacman::p_load(tidyverse)
set.seed(17)
n <- 10
mu <- 4
X <- rnorm(n, mean = mu, sd = 1)
print(X)
```

The sample mean is `r round(mean(X),3)`. The question is whether values at least this extreme arise plausibly from a population with $\mu = 4$. Walking through the NHST procedure:

1. The null hypothesis is that the population mean equals the theoretical value (here the scale midpoint, 4), i.e., $\mu = 4$. The alternative is $\mu \neq 4$.
2. The test statistic is the *T* statistic — the same one used for interval estimation when the population variance is unknown.
3. The decision threshold is the conventional 5%.

The test-statistic computation and the decision are handled in one call to R's `t.test()`:

```{r,Os.ttest}
result <- t.test(X, mu = mu)
print(result)
```

The realised test statistic is `r round(result$statistic,3)`, and the probability of obtaining a value this extreme or more so from the $t$ distribution with `r result$parameter` degrees of freedom is `r round(result$p.value,3)`. This exceeds 5%, so the result is not unusually rare: drawing a sample mean of `r round(mean(X),3)` from a normal population with mean 4 is not surprising, and there is no statistically significant difference.

In a report, the result is conventionally written as "$t(9) = 0.66776, p = 0.5151, \text{n.s.}$", where "n.s." stands for *not significant*.

This example may feel circular: we drew from a normal with mean 4 and then concluded that the mean is not different from 4. But consider the next example:

```{r,Os.ttest2}
n <- 3
mu <- 4
X <- rnorm(n, mean = mu, sd = 1)
mean(X) %>%
  round(3) %>%
  print()
result <- t.test(X, mu = mu)
print(result)
```

With $n = 3$ and a sample mean of `r round(mean(X),3)`, the $t$ statistic exceeds the 5% critical value. The conclusion is that the observed value is too extreme to have come from a population with mean 4, so we reject the null. The simulation used $\mu = 4$, but a small sample drawn from that population can wander far from the population mean.

## The two-sample test

Now consider the two-sample test, used to compare the mean of an experimental group against that of a control group, capitalising on random assignment to estimate the average causal effect. The null is "no group difference"; the alternative is its negation. Assuming samples from normal populations, the test statistic again follows a $t$-distribution.

Walking through the procedure:

1. The null is "the two population means are equal." Writing the two population means as $\mu_1$ and $\mu_2$, this is $\mu_1 = \mu_2$, or equivalently $\mu_1 - \mu_2 = 0$. The alternative is $\mu_1 \neq \mu_2$.
2. The test statistic is the $T$ from the unknown-variance case.
3. The decision threshold is 5%.

To check this, generate synthetic data. Let `n1` and `n2` be the per-group sample sizes (here both equal to 10 for simplicity). Set $\mu_1$ as the population mean of group 1 and write $\mu_2 = \mu_1 + \delta$; with $\delta = 0$ the means are equal, and with $\delta \neq 0$ they differ. Finally, set the population SD for both groups.

The test asks whether the observed difference $d$ could plausibly have arisen from a population in which the mean difference is zero. The test statistic $T$ is

$$ T = \frac{d - \mu_0}{\sqrt{U^2_p \cdot \frac{n_1 + n_2}{n_1 n_2}}}$$

where $d$ is the difference of the two sample means and $U^2_p$ is the **pooled unbiased variance** — an estimate of the common population variance pooled across both groups. With $S^2_1$ and $S^2_2$ as the two sample variances,

$$ U^2_p = \frac{n_1 S^2_1 + n_2 S^2_2}{n_1 + n_2 - 2}. $$

In essence: weight each sample variance by its sample size, then divide by $n_1 + n_2 - 2$ to obtain an unbiased pooled estimate.

A concrete numerical example. Generate the data, inspect the sample means, then run `t.test()`:

```{r,Ts.ttest}
n1 <- 10
n2 <- 10
mu1 <- 4
sigma <- 1
delta <- 1
mu2 <- mu1 + (sigma * delta)

set.seed(42)
X1 <- rnorm(n1, mean = mu1, sd = sigma)
X2 <- rnorm(n2, mean = mu2, sd = sigma)

X1 %>%
  mean() %>%
  round(3) %>%
  print()
X2 %>%
  mean() %>%
  round(3) %>%
  print()

result <- t.test(X1, X2, var.equal = TRUE)
print(result)
```

We set $\mu_1 = 4$ and $\mu_2 = 4 + 1$, but the sample means are `r round(mean(X1),3)` and `r round(mean(X2),3)` — not strikingly different in this draw. The $t$ statistic is `r abs(result$statistic)`, with $p = $ `r result$p.value` on `r result$parameter` degrees of freedom; above the 5% threshold, so we fail to reject the null.

By construction the population means *do* differ ($4 \neq 4 + 1$); this is a misjudgement, a **Type II error**. In actual research we do not know the population means, so we cannot tell whether such an error has occurred.

In the example above we created two separate objects `X1` and `X2` to make the two groups explicit. In practice the grouping is usually a variable inside a data frame, and the test is written with R's `formula` interface:

```{r}
dataSet <- data.frame(group = c(rep(1, n1), rep(2, n2)), value = c(X1, X2)) %>%
  mutate(group = as.factor(group))
t.test(value ~ group, data = dataSet, var.equal = TRUE)
```


## The two-sample test (Welch's correction)

The call to `t.test()` above used the option `var.equal = TRUE`, assuming the two groups share a common variance. The original $t$-test does assume equal variances, but in practice this assumption cannot be taken for granted. The standard preliminary check is Levene's test for equality of variances; in R, the `car` and `lawstat` packages provide implementations. Below we use `leveneTest()` from `car`:

```{r}
#| message: FALSE
pacman::p_load(car)
leveneTest(value ~ group, data = dataSet, center = mean)
```

The $p$-value is too large to reject the null of equal variances, so we may proceed under the equal-variance assumption. If Levene's test were significant, the equal-variance assumption would have to be dropped. Doing so in `t.test()` is trivial — set `var.equal = FALSE`:

```{r}
result2 <- t.test(value ~ group, data = dataSet, var.equal = FALSE)
print(result2)
```
The output title now reads "Welch Two Sample t-test" — the $t$-test with **Welch's correction**. Note also that the degrees of freedom are now non-integer (`r round(result2$parameter,3)`). Welch's correction adjusts the degrees of freedom to compensate for unequal variances. A report would write "$t($ `r round(result2$parameter,3)` $)=$ `r round(result2$statistic,3)`, $p=$ `r round(result2$p.value,3)`"; the non-integer degrees of freedom signal that Welch's correction has been applied.

Equality of variances is a *special case* of the more general (unequal-variance) scenario, so it is reasonable to run Welch's test from the start. R's `t.test()` defaults to `var.equal = FALSE` accordingly: unless told otherwise it does not assume equal variances. This is preferable because it avoids the multiplicity of running an auxiliary equal-variance test.

### Effect sizes

In the example above the synthetic data had $\mu_1 = 4$ and $\mu_2 = \mu_1 + \sigma d$, so $\mu_1 \neq \mu_2$ by construction — yet we failed to detect the difference. Statistical significance is a *statistical* matter; what we usually want to know in substantive research is whether there is a real, substantive difference. Making "obtain a significant result" the goal of an analysis is therefore plainly the wrong objective.[^8.1]

[^8.1]: In, say, physics, measurement precision is high and one is examining a single physical world: there is no need to reason probabilistically about whether predictions are true or false. In such a setting — let us call it a world of clear theoretical truth — information about statistical significance is at most a supplementary device for supporting a theory, and reporting test results is largely a rhetorical convention for writing papers. The verification of Newton's laws or Einstein's relativity hinges on whether measured values agree with theoretical predictions within the measurement error; statistical significance is, at best, a secondary concern. Psychology, dealing with small human samples, has little choice but to rely on statistical adjudication. But that should not let us forget that the substantive difference is what really matters.

When can we say firmly that there is a difference? The four panels of distributions below clarify the answer:


```{r}
#| echo: FALSE
#| message: FALSE
pacman::p_load(tidyverse)
pacman::p_load(ggplot2)
pacman::p_load(gridExtra)
dL <- 2
dS <- 0.2
sL <- 2
sS <- 0.2

g1 <- ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = dL, sd = sS), color = "red") +
  stat_function(fun = dnorm, args = list(mean = -dL, sd = sS), color = "blue") +
  xlab("") +
  ylab("") +
  theme(axis.text.y = element_blank())
g2 <- ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = dS, sd = sS), color = "red") +
  stat_function(fun = dnorm, args = list(mean = -dS, sd = sS), color = "blue") +
  xlab("") +
  ylab("") +
  theme(axis.text.y = element_blank())
g3 <- ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = dL, sd = sL), color = "red") +
  stat_function(fun = dnorm, args = list(mean = -dL, sd = sL), color = "blue") +
  xlab("") +
  ylab("") +
  theme(axis.text.y = element_blank())
g4 <- ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = dS, sd = sL), color = "red") +
  stat_function(fun = dnorm, args = list(mean = -dS, sd = sL), color = "blue") +
  xlab("") +
  ylab("") +
  theme(axis.text.y = element_blank())
g <- grid.arrange(g1, g2, g3, g4)
```

The left column shows pairs with a large mean difference, the right column small. The top row shows small variances, the bottom row large. In which of the four panels is "the two groups differ" most defensible?

The top-left case is the clearest: the two groups are obviously separated, with negligible overlap. The bottom-left case has the same mean difference but much wider within-group spread, so the groups overlap and many individuals fall in the region the verdict "different" would not apply to. The top-right case has minimal overlap but a small mean difference, so it is borderline. The bottom-right case has both a small difference and large overlap; a verdict of "different" would have many exceptions. Take the claim that "men are stronger than women" (a difference in body strength on average): even if true on average, there are surely many men weaker than many women. When the exceptions are this prevalent, a statistically significant difference would not feel meaningful.

The point: judging whether groups differ depends not just on the mean difference but also on the variance. The **standardised difference** — the mean difference divided by the standard deviation — is therefore central. This is the **effect size**.[^8.2]

[^8.2]: Substantive difference is more meaningful than effect size, and effect size more meaningful than statistical significance. See @Toyoda2009 for an extended treatment.

The data above used $\mu_2 - \mu_1 = \sigma d$; the effect size for a mean difference is

$$ es = \frac{\mu_1 - \mu_2}{\sigma}, $$

so the $d$ in the simulation was exactly the effect size. In real data the population mean and SD are unknown and the effect size must be estimated from the sample. R's `effsize` package provides routines for this:

```{r}
pacman::p_load(effsize)
cohen.d(value ~ group, data = dataSet)
cohen.d(value ~ group, data = dataSet, hedges.correction = TRUE)
```

After a mean-difference test it is standard practice to also report an effect size such as Cohen's $d$ or Hedges's $g$.

## The paired-samples test

When the two groups are not independent — as in a pre/post within-subjects design — the formulation of the $t$-test changes. For independent samples we considered the distribution of the difference of group means, $\mu_1 - \mu_2$. For paired samples we consider the per-individual difference $X_{i1} - X_{i2} = D_i$. Testing this one statistic makes the paired test a special case of the one-sample test. The standard error of $D$ is estimated as $U_D / \sqrt{n}$, where $U_D$ is the sample SD of the differences.[^8.3] The test statistic $T$ is

$$ T = \frac{\bar{D}}{U_D/\sqrt{n}} = \frac{\sum D_i/n}{\sqrt{\frac{\frac{1}{n-1}\sum(D_i-\bar{D})^2}{n}}}.$$

[^8.3]: Since the design is paired, $n$ is the same on both occasions.

In `t.test()`, set `paired = TRUE`.

### Constructing synthetic data

Two approaches are common for generating paired-sample synthetic data. The first models a generative process directly:
```{r}
n <- 10
mu1 <- 4
sigma <- 1
d <- 1
X1 <- rnorm(n, mu1, sigma)
X2 <- X1 + sigma * d + rnorm(n, 0, sigma)
t.test(X1, X2, paired = TRUE)
```

Group 1 is sampled around its mean $\mu_1$; group 2 is obtained by adding a constant effect $\sigma d$ to each value of group 1, plus additional measurement noise. This mirrors a concrete data-generation story, but the noise is in some sense double-counted.

The second approach generalises beyond pre/post designs to any setting that has "pairing." Pairing means the two values are drawn jointly from a **bivariate** normal — i.e., each variable is normally distributed and the two are correlated. The univariate normal is
$$X \sim N(\mu, \sigma),$$
while the multivariate normal — generating several variables jointly — is written
$$ \mathbf{X} \sim MVN(\mathbf{\mu}, \mathbf{\Sigma}),$$

where $\mathbf{X}$ and $\mathbf{\mu}$ are $n$-vectors and $\mathbf{\Sigma}$ is the variance–covariance matrix. In the bivariate case,

$$\mathbf{\Sigma} = \begin{pmatrix} \sigma_1^2 & \sigma_{12}\\ \sigma_{21} & \sigma_2^2 \end{pmatrix} = \begin{pmatrix} \sigma_1^2 & \rho_{12}\sigma_1\sigma_2 \\ \rho_{21}\sigma_2\sigma_1 & \sigma_2^2 \end{pmatrix}.$$

The off-diagonal covariance $\sigma_{ij}$ can be written in terms of the correlation $\rho_{ij}$; thus the data-generation process explicitly encodes a between-variable correlation. Synthetic data on this model:

```{r}
#| message: FALSE
pacman::p_load(MASS) # for multivariate normal random numbers
n <- 10
mu1 <- 4
sigma <- 1
d <- 1
mu <- c(mu1, mu1 + sigma * d)
rho <- 0.4
SIG <- matrix(c(sigma^2, rho * sigma * sigma, rho * sigma * sigma, sigma^2), ncol = 2, nrow = 2)
X <- mvrnorm(n, mu, SIG)
t.test(X[, 1], X[, 2], paired = TRUE)
```

Effect sizes are computed as for the independent-samples case:

```{r}
cohen.d(X[, 1], X[, 2])
cohen.d(X[, 1], X[, 2], hedges.correction = TRUE)
```

### Directionality

So far we have used hypotheses of the form "is there a difference?" The question of whether the difference is in the positive or negative direction has not been at issue; the test statistic's distribution has been used with the *two tails* combined to define the significance level.

In a pre/post experiment, however, the *direction* — has performance improved, or deteriorated? — is often the substantive question. An effect that is the wrong sign may be no use. For a directional hypothesis only one tail of the test statistic's distribution is relevant. In `t.test()`, the `alternative` option encodes this.

Setting `t.test(x, y, alternative = "less")` tests the null against $x < y$; `alternative = "greater"` tests against $x > y$. The default is `alternative = "two.sided"` — a two-tailed test.

One subtlety: moving from two-tailed to one-tailed *lowers* the critical value the test statistic must exceed. A one-tailed test is therefore a *less conservative* test. Some take the view that defaulting to two-tailed is sufficiently strict to be safe; the better practice is to use the null hypothesis appropriate to the research question.

## Exercises

1. Draw 30 samples from a normal with mean 50 and SD 10, and test whether the sample mean differs from the population mean. Report the result using the conventions of the Japanese Psychological Association's *Manual for the Preparation and Submission of Papers*.
2. Use the data below to test whether the means of two independent groups differ. Report the result in the same convention.
$$ group1 = \{45, 50, 55, 60, 65\} $$
$$ group2 = \{57, 60, 62, 77, 75\} $$

3. Practise the paired-samples $t$-test using bivariate-normal synthetic data. With $n = 20$, mean vector $\mu = (12, 15)$, and covariance matrix $\Sigma = \begin{pmatrix} 4 & 2.8 \\ 2.8 & 4\end{pmatrix}$, generate the sample and run the paired test. Report the result in the same convention.

4. Plot the $t$-density for $\mathrm{df} = 10, 20, 30$ alongside the standard-normal density. How does the $t$-distribution change as the degrees of freedom increase?
5. For a $t$-distribution with 15 degrees of freedom, compute the critical values (theoretical thresholds for the decision) at the 5% level for one-tailed and two-tailed tests.

```{r}
#| echo: FALSE
#| eval: FALSE
#| include: FALSE

pacman::p_load(tidyverse)

# Exercise 1
set.seed(123)
sample_data <- rnorm(30, mean = 50, sd = 10)

# run the t-test
t_test_result <- t.test(sample_data, mu = 50)

t_test_result

# write up
cat("t(", t_test_result$parameter, ") = ", t_test_result$statistic, ", p = ", t_test_result$p.value, sep = "")

# A one-sample t-test of 30 draws from N(50, 10) against mu = 50 yields no significant difference
# (t(29) = -0.195, p = 0.847).

# Exercise 2
group1 <- c(45, 50, 55, 60, 65)
group2 <- c(57, 60, 62, 77, 75)

t_test_result <- t.test(group1, group2, var.equal = TRUE)
t_test_result

cat("t(", t_test_result$parameter, ") = ", t_test_result$statistic, ", p = ", t_test_result$p.value, sep = "")

# The means of the two independent groups differ significantly
# (t(8) = -2.838, p = 0.022).

# Exercise 3
pacman::p_load(MASS)

set.seed(123)
n <- 20
mu <- c(12, 15)
Sigma <- matrix(c(4, 2.8, 2.8, 4), ncol = 2)
data <- mvrnorm(n, mu = mu, Sigma = Sigma)

t_test_result <- t.test(data[, 1], data[, 2], paired = TRUE)
t_test_result

cat("t(", t_test_result$parameter, ") = ", t_test_result$statistic, ", p = ", t_test_result$p.value, sep = "")

# A paired-samples t-test indicates a significant difference
# (t(19) = -9.256, p < 0.001).

# Exercise 4
x <- seq(-4, 4, length = 1000)
df_values <- c(10, 20, 30)
norm_density <- dnorm(x)

df <- data.frame(x = x, norm = norm_density)
for (df_val in df_values) {
  df[paste("t_df", df_val, sep = "_")] <- dt(x, df = df_val)
}

df_long <- gather(df, key = "distribution", value = "density", -x)

ggplot(df_long, aes(x = x, y = density, color = distribution)) +
  geom_line(size = 1) +
  labs(
    title = "t-distributions with different degrees of freedom vs the standard normal",
    x = "x", y = "Density"
  ) +
  scale_color_manual(
    values = c("black", "blue", "green", "red"),
    labels = c("Normal", "t (df = 10)", "t (df = 20)", "t (df = 30)")
  ) +
  theme_classic()

# Exercise 5
critical_value_one_sided <- qt(0.95, df = 15)
critical_value_two_sided <- qt(0.975, df = 15)

cat("One-tailed critical value: ", critical_value_one_sided, "\n")
cat("Two-tailed critical value: ", critical_value_two_sided, "\n")
```
